\Large{Adversarial Training against Systematic Uncertainty}

### 2016-10-27 tex/paper.tex
edit physics toy

In collisions at the Large Hadron Collider, massive particles can be produced at
such high energies that their decay products are collimated and the resulting
jets overlap. Still, deducing whether an observed jet is either due  to a single
hadronic particle or to the decay of a massive object into multiple hadronic
particles  is critical for understanding the nature of the particles produced in
the collision \citep{butterworth2008jet,baldi2016jet}. Critically, this
classification problem is often made difficult by the presence of simultaneous
interactions, known as pileup, which contributes significant energy depositions
unrelated to the particles of interest.

In this setting, we reuse the classification problem of \citep{baldi2016jet}
between single jets produced in quark or gluon fragmentation and two overlapping
jets produced when a high-velocity $W$ boson decays to a collimated pair of
quarks. As nuisance parameter, we consider events without pileup ($Z=0$) and
events with pileup ($Z=1$), for which an average number of $\langle \mu \rangle =
50$ unrelated interactions are overlaid. Our goal is to build an accurate
classifier, for which we also want to minimize the effects due to the
uncertainties on the nuisance. More specifically, we choose to recast the
classification problem as a hypothesis test between signal+background (jets
originating from $W$ bosons or from single quarks and gluons) and background
only (jets originating from single quarks and gluons only) and tune the decision
threshold of our classifier by maximizing its approximate median significance
(AMS), when uncertainties in the background are taken into account (see Eqn.~20
of \citep{adam2014higgs}). Our motivation is that reducing the effects of
uncertainties by requiring independence of $Z$ with the classifier output $f(X;\theta_f)$ should
allow for a larger maximum significance.

To minimize the effects of $Z$ in the background events, we train a classifier
using  Algorithm~\ref{alg:adversarial-training} but consider the adversarial
term ${\cal L}_r$ conditioned on $Y=0$ only. Both $f$ and $r$ are neural
networks with 3 hidden layers of 64 nodes and ReLU activations, each terminated
by a single final output node with a sigmoid activation. Experiments are performed
on the high-level features data described in \citep{baldi2016jet}, on a subset
of 150000 samples for training while AMS is evaluated on an independent test set of 5000000
samples. Results reported below are averages over 5 runs.

>>>>

Jets, or collimated sprays of particles produced
by high energy quarks and gluons, are produced ubiqitiously at high
energy colliders like the Large Hadron Collider \citep{LHCMachine}.  In a wide
array of new physcis models, new highly massive particles can decay
though known heavy particles in Standard Model (SM), like the $W$, $Z$, Higgs bosons, or top
quarks which subsequently decay to mutiple quarks.  When these these
heavy SM particles have significant energy, the Lorentz boost causes
their quark decay products to merge into a single jet which has a rich
internal substructure (see e.g.  \citep{Altheimer:2012mn,Altheimer:2013yza} and references
therin).  Distinguishing these \textit{boosted jets} produced by heavy SM particles
from vanilla quarks and  gluons is a fundamental challenge in
searching for signs of new high mass particles.

Challenging in it's own right, this classification problem is
made all the more difficult by the presence of pileup, or multiple simultaneous
proton-proton interactions occuring at the same time as the
primary interaction.  These pileup interactions produce additional
particles that can contributie singificant energies to jets unrelated to
the underlying discrimnating information. Thus pileup acts as a source of noise in
the classification problem.

% In collisions at the Large Hadron Collider, massive particles can be produced at
% such high energies that their decay products are collimated and the resulting
% jets overlap. Still, deducing whether an observed jet is either due  to a single
% hadronic particle or to the decay of a massive object into multiple hadronic
% particles  is critical for understanding the nature of the particles produced in
% the collision \citep{butterworth2008jet,guest2016jet}. Critically, this
% classification problem is often made difficult by the presence of simultaneous
% interactions, known as pileup, which contributes significant energy depositions
% unrelated to the particles of interest.

The classification challenge used here is common in jet substructure
studies (see \citep{Khachatryan:2014vla,ATL-PHYS-PUB-2015-033,wbosonATLAS} and
references therein): we aim to discriminate between jets produced by quark or gluon fragmentation (the
background) and  boosted $W$ boson jets containing a collimated pair of
quarks (the signal).  We reuse the datasets used in reference
\citep{guest2016jet}.  Briefly summarizing, jets are built with the
anti-$k_T$ algoritm \citep{Cacciari:2008gp} 
with radius parameter $R=1.2$, and trimmed \citep{Krohn:2009th} with subjets built with
the $k_T$ algorithm and parameter $f_{cut}=0.2$. The features used
in the classifiers are: trimmed jet invariant mass, N-subjettiness
$\tau_{21}^{\beta=1}$ \citep{nsub,Thaler2012}, and the energy correlation
functions \citep{Larkoski2013} $C_{2}^{\beta=1}$, $C_{2}^{\beta=2}$,
$D_{2}^{\beta=1}$, $D_{2}^{\beta=2}$.

As nuisance parameter, we consider events without pileup ($Z=0$) and
events with pileup ($Z=1$), for which an average number of $\langle \mu \rangle =
50$ unrelated additional pileup interactions are overlaid. Our goal is to build an accurate
classifier, for which we also want to minimize the effects due to the
uncertainties on the nuisance. More specifically, we choose to recast the
classification problem as a hypothesis test between signal+background (jets
originating from $W$ bosons or from quarks and gluons) and background
only (jets originating from quarks and gluons only) and tune the decision
threshold of our classifier by maximizing its approximate median significance
(AMS), when uncertainties in the background are taken into account (see Eqn.~20
of \citep{adam2014higgs}). Our motivation is that reducing the effects of
uncertainties by requiring independence of $Z$ with the classifier output $f(X;\theta_f)$ should
allow for a larger maximum significance.

To minimize the effects of $Z$ in the background events, we train a classifier
using  Algorithm~\ref{alg:adversarial-training} but consider the adversarial
term ${\cal L}_r$ conditioned on $Y=0$ only. Both $f$ and $r$ are neural
networks with 3 hidden layers of 64 nodes and ReLU activations, each terminated
by a single final output node with a sigmoid activation. Experiments are performed
on a subset of 150000 samples for training while AMS is evaluated on an independent test set of 5000000
samples. Both training and testing samples are weighted such
that the signal-to-background ratio is 1/10, allowing us to probe the
efficacy of the method proposed here in a background dominated
environment. Results reported below are averages over 5 runs.
